---
title: "Re-create a table in gt()"
author: "Kasia Banas"
date: "2024-10-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

We will use the package `gt` to create a table with data about stroke mortality in Scotland. If you haven't installed this package yet, please do it now.

We'll be working with a stroke mortality dataset from Public Health Scotland. Follow the link below to find the data dictionary, and a downloadable csv file. https://www.opendata.nhs.scot/dataset/scottish-stroke-statistics/resource/19c01b59-6cf7-42a9-876a-b07b9b92d6eb

You will also need a dataset with health board names, you can find it here: https://www.opendata.nhs.scot/dataset/geography-codes-and-labels/resource/652ff726-e676-4a20-abda-435b98dd7bdc

## Load the libraries
```{r, libraries}
library(tidyverse)  
library(gt) 
library(glue) # This package can be used for the subtitle of the table
```

## Import the data
```{r, import data}
# Read in the data

stroke_mortality_raw <- read_csv("https://www.opendata.nhs.scot/dataset/f5dcf382-e6ca-49f6-b807-4f9cc29555bc/resource/19c01b59-6cf7-42a9-876a-b07b9b92d6eb/download/stroke_mortalitybyhbr.csv")
 
hb <- read_csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv")
 
```

## Join the datasets
Join the two datasets to get nice health board names.

```{r}
joined_data <- hb %>% 
  full_join(stroke_mortality_raw, by = c("HB" = "HBR"))
```


## Build an MVP
Create a first version of the table, using only tidyverse functions. Pipe it to `gt()` to see how it is formatted.

```{r}
table_data <- joined_data %>% 
  filter(Year == 2018,
         AgeGroup == "75plus years",
         Sex != "All",
         HBName %in% c("NHS Lothian", "NHS Greater Glasgow and Clyde")) %>% 
  select(Diagnosis, Sex, NumberOfDeaths, CrudeRate, EASR, HBName)
gt(table_data)
```


## Improve
Add `gt()` functions to get your table closer and closer to the end product. 
If you need some hints, scroll down this document. 

```{r}
table_data <- joined_data %>% 
  filter(Year == 2018,
         AgeGroup == "75plus years",
         Sex != "All",
         HBName %in% c("NHS Lothian", "NHS Greater Glasgow and Clyde")) %>% 
  select(Diagnosis, Sex, NumberOfDeaths, CrudeRate, EASR, HBName) %>% 
  group_by(HBName)
gt(table_data) %>% 
  cols_label(NumberOfDeaths = "Number of Deaths", 
             CrudeRate = "Crude Rate",
             EASR = "European Age-Sex Standardised Rate") %>% 
  cols_align(align = "center", 
             columns = NumberOfDeaths) %>% 
  fmt_number(columns = c(CrudeRate, EASR), decimals = 2) %>% 
  tab_header(title = "Stroke Mortality in 2018 in Adults Aged 75+",
             subtitle = "Data from NHS Lothian and NHS Greater Glasgow and Clyde") %>%
  tab_spanner(label = "Rate per 100k population",
            columns = c(CrudeRate, EASR)) %>% 
  summary_rows(columns = c(NumberOfDeaths, CrudeRate, EASR), 
               fns = list(Average = ~mean(., na.rm = TRUE)),
               fmt = list(~ fmt_number(., decimals = 2))) %>% 
  grand_summary_rows(columns = c(NumberOfDeaths, CrudeRate, EASR), 
               fns = list("Ovarall Average" = ~mean(., na.rm = TRUE)),
               fmt = list(~ fmt_number(., decimals = 2)))
```



















## Hints

Here are the gt() functions that you're likely to need:
- cols_move()
- cols_label()
- cols_align()
- summary_rows()
- grand_summary_rows()
- fmt_number()
- tab_header()
- tab_spanner()





